Concave reduction of swarm voids

### 2018-06-03 paper.tex
IEEE Access modifications

\begin{abstract}
%\boldmath
Swarms consist of many agents and this is exploited to make them tolerant of failures. Failures occur for may reasons; for example: resource exhaustion, component failure, or a disruption from an external event. The loss of agents reduces the size of a swarm and changes the structure in places where the agents are lost. A swarm's structure can also be irregular from an initial deployment or due to an obstacle in its path.

These changes in the structure and/or size do not stop a swarm from functioning but can affect the outcome of a task to which the swarm is being applied.

The principle behind concave reduction is to counter the effect of agent loss or irregular structures within a swarm. Depending upon the type of failure or disruption sections of the swarm can develop holes (areas that should contain agents) or the overall shape may become distorted. Concave reduction is a technique where each agent identifies its position in relation to its neighbours and if the agent meets specific criteria it will attempt to move in such a way as to remove the gap between two of its neighbours. These changes in direction of multiple agents have a restructuring effect which results in a more uniform distribution of agents and reduces the perimeter size for the number of agents in the swarm.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

\section{Swarm coordination - concave reduction}\label{chapter:ConcaveReduction}
When cohesion and repulsion (\textit{inter-agent vectors}) are the only field effects creating a swarming effect the number of stable structures that can develop is limited. These structures are either straight edges or partial lattices. Partial lattices can create concave anomalies (dents) and convex anomalies (peaks) in a perimeter~(Figure~\ref{concave:SwarmStableShape1}). An anomaly is any construct within the swarm that is not a hexagonal lattice or an agent distribution that causes the swarm to deviate from having only convex or flat edges.

@@

When discussing concave reduction the concave angle of interest is the relationship between an agent and two of its adjacent neighbours such that the neighbours have no visibility of each other and the neighbour-agent-neighbour relationship produces an angle~$< 180^\circ$. This is unrelated to the aggregate angle used to identify the perimeter type that the agent is a member of. Concave reduction therefore affects both inner~(Figure~\ref{fig:InnerPerimeter1}) and outer~(Figure~\ref{fig:OuterPerimeter1}) perimeter agents. Figures~\ref{fig:OuterPerimeter1} and \ref{fig:InnerPerimeter1} show candidate agents identified by this process in black.

@@

The purpose of concave reduction is to alter the overall structure of the swarm by replacing the calculated \textit{movement-direction vector} on anomalous perimeter agents. These perimeter based changes cause a cascading movement within the swarm as non-perimeter based agent's \textit{interaction vectors} attempting to move towards a position of equilibrium to their neighbours. The movement of the concave-perimeter~(internal void) agents creates a tendency for the void to be removed~(Figures~\ref{fig:InnerPerimeter2}~and~\ref{fig:InnerPerimeter3}) by percolating the void out of the swarm. On the outer perimeter deformities (indents) are straightened~(Figure~\ref{fig:OuterPerimeter3}) and eventually a perimeter agent and its neighbours will tend towards creating a convex edge.

@@

\subsection{Full-perimeter coordinator detection}\label{sec:PerimeterAgentDetection} 
The process of detecting a perimeter agent involves several steps. Primarily it involves identifying agents that are not surrounded by interconnected neighbours~(Figure~\ref{reduced:Perimeter1}). The processes are discussed in more detail below:

@@

The first check is a short circuit count of the number of neighbours. If an agent has $\leq 4$ neighbours and the swarm parameters are for a hexagonal swarm configuration then the agent is a perimeter agent and therefore a coordinator. This identification process is shown in~Figure~\ref{algo:SmallNeighbourCount}.

@@

If the short circuit neighbour count check fails then further conditions must be checked. The additional checks require a dictionary of agent/angular values to be generated~\eqref{eq:ConcaveVoidPhysics1}. Figure~\ref{algo:getNeighbourAngles} shows how the dictionary of agent/angle values are produced.

@@

Figure \ref{algo:getNeighbourAngles} shows the logic to produce the sorted neighbour/angle dictionary. $sort(.)$ sorts the dictionary set generated in~\eqref{eq:ConcaveVoidPhysics1} by ascending angle to produce a dictionary of neighbour agents with their relative angle to the first detected neighbour as shown in~Figure~\ref{reduced:Perimeter1}.

@@

Using the angular dictionary from Figure~\ref{algo:getNeighbourAngles} it is possible to identify if adjacent neighbours can detect each other. If the agent is within all of its neighbours boundary then it is not on the perimeter~(Figure~\ref{algo:CheckVisibility}).

@@

The full visibility check~(Figure~\ref{algo:CheckVisibility}) takes each pair in turn and using the cosine rule~\eqref{eq:CosineRule1} checks for a `visibility gap'.

@@

There is one exception to the neighbour visibility check. Due to compression of the swarm, which can be caused by an initial deployment configuration or when an obstacle is in the path of the swarm, the agent's neighbours are able to `see' other neighbours but a pair of sequential neighbours could create an angle $> 180^\circ$ as shown in~Figure~\ref{reduced:Perimeter2}. In this case the agent is on the outside of the enclosed neighbour space, the agent is therefore on a perimeter.

@@

This phenomenon requires the detection process to identify the angles that all the agents create from an arbitrary point, in this case the first identified neighbour. The neighbours must then be checked to determine if any neighbour pair create an angle $> 180^\circ$~(Figure~\ref{reduced:Perimeter1}~Figure~\ref{algo:CheckConvex}). This process includes a short circuit in that as soon as a gap is found the check terminates.

@@

The complete perimeter detection algorithm using this `cyclic-angular-neighbour-check' methodology is show in~Figure~\ref{algo:CheckPerimeter}. The first part of the algorithm is the short circuit check~(\textit{SmallNeighbourCount(b)}). This is only followed by the visibility check (\textit{CheckVisibility(b,angles)}) if the short circuit fails. If the visibility check is confirmed, which is another short circuit method, the convex check is carried out~(\textit{CheckConvex(angles)}). This sequence ensures the minimum computational overhead for the agent check.

@@

In a static swarm, where there are no \textit{destination vectors}, concave reduction will result in a restructuring motion that creates a more `rounded' swarm~(Figure~\ref{fig:OuterPerimeter3}). Concave reduction also creates a surround effect as it removes voids from a swarm~(Section~\ref{voids:ObjectSurrounding}). When concave reduction is applied to a mobile swarm the void closing effect can be used to improve reconnaissance coverage as a swarm passes around obstacles~(Section~\ref{concave:mobileSwarm1}).

@@

The implementation of concave reduction identification is incorporated into the perimeter detection algorithm as shown in Figure~\ref{algo:checkVisibility2}.\\

@@

\section{Concave reduction agent movement}\label{concave:AgentMovement}
Adding a further characteristic to the motion of a swarm necessitates a revision to the existing agent model~\eqref{eq:InterAgentMovement2}. With concave reduction this revision is based upon the identification of the concave perimeter edges as shown in~(Figure~\ref{concave:VoidConcave1}) $(b^{'},b,b^{''})$.

@@

When an agent is identified as being a component of a concave characteristic~(Figure~\ref{algo:checkVisibility2}) the normal \textit{movement-direction vector}~\eqref{eq:InterAgentMovement2} is replaced by a \textit{concave reduction vector}. This new vector causes the agent to move in a direction that will reduce or remove a concave edge by moving the agent towards the identified gap. The effect of this will be to either straighten an outer perimeter or reduce/remove a void. This change in direction increases the distance and magnitude variances (jitter) due to the changes it will induce in the neighbours \textit{interaction vectors}.

Figures~\ref{fig:InterAgentEffect1} and \ref{fig:InterAgentEffect2} shows this effect in more detail. Figure~\ref{fig:InterAgentEffect1}~shows the initial positions of the agents before the concave reduction affects the agent. Figure~\ref{fig:InterAgentEffect2}~shows the positive and negative effect on the inter-agent distances that the movement creates. The aggregate change is an increase in the inter-agent distances and an increase in the resultant magnitude effect.

@@

\subsection{Perimeter Exceptions}\label{concave:Exceptions}
When a swarm is forming or disrupted through either agent failures, catastrophic events or interactions with obstacles sections of the swarm may become isolated. When this occurs the sections can be considered as being a separate sub-swarms. It is also possible for sections of the swarm to remain connected through tenuous links of either 1 or 2 agents as shown in Figures~\ref{fig:Connector1} and \ref{fig:Connector2}. In the case of a link connection~(Figure~\ref{fig:Connector1}) concave reduction can be applied as normal. In the case of a `bridge' link (Figure~\ref{fig:Connector2}) there is an issue in that there is more than one concave gap. As there is no communications infrastructure in swarm models there are limited options for handling this situation. One option is allow the swarming algorithm to apply as usual and no concave reduction takes place; the reasoning being that the overall structure of the swarm is unknown therefore the movement cannot be optimised. An alternative would be to select either the largest or the smallest gap; again as the structure of the swarm is unknown this may not be beneficial. A third option would be to select one of the gaps in some way (randomly, first seen, last seen) and implement the concave reduction at that point. The approach taken in this paper is to select the first gap that is detected. Selecting the first gap reduces the computational overhead of the algorithm.

@@

Figure~\ref{concave:BaselineConcaveEffectDist} shows the changes in the inter-agent distances that result from the concave reduction. The experiment demonstrates that with concave reduction the average distance increases and the variance of the distances increases. This is due to the `pulling' effect of the \textit{concave reduction vector} on the perimeter. The algorithm distorts the distribution of the agents as discussed in section~\ref{concave:AgentMovement}. The initial expansion of the swarm is similar for the baseline and the concave reduction however at 9 seconds into the simulation the \textit{concave reduction vectors} affects the swarm sufficiently to prevent the average distance reducing to the same level as the baseline. Once the swarm has stabilised with the concave reduction enabled the average distance and variance remain constant but above the baseline.

@@

Figure~\ref{concave:BaselineConcaveEffectMag} shows the change in the \textit{inter-agent vector magnitudes} that the concave reduction introduces. Both the average magnitude increases and the variance. The magnitude increases as the \textit{concave reduction vector} has caused the agents to move outwards increasing the area of the swarm. The increase in cohesion is a direct result of the increased distance. The field effects still intersect but due to the average distance of the agents being further apart the repulsion is reduced. The agents being further apart causes the increase in cohesion as the algorithm attempts to prevents the swarm from breaking up. The variance increase is caused by the concave reduction effect moving selected agents into less optimal positions.

@@

Figure~\ref{concave:BaselineConcavePerimeter} shows the effect on the number of perimeter agents. Initially the perimeter size is minimally affected due to the swarms compression however after the expansion phase of the swarm (8 seconds) more perimeter agents are identified as concave edges are formed. As these concave edges are `straightened' they create further anomalies that ripple through the swarm. This rippling is identified by the erratic changes in the number of perimeter agents. After approximately 30 seconds the swarm has been forced into a less angular structure with curved edges and the number of perimeter agents falls below the minimum of the baseline. The shape of the swarm is still undergoing change following this and the perimeter count continues to fluctuate.

@@

The effect on the structure of the swarm caused by the concave perimeter agents moving towards the gap agents is to pull the internal agents forward. This pulling causes the swarm to develop a more rounded structure but the effect also have a negative impact. If the tolerance of the agent's movements (the difference between the agent ranges for repulsion and cohesion) is too small then the distortion effect of the concave reduction can create additional voids. The perimeter agents will then move so as to remove the defect~(Figure~\ref{fig:OuterPerimeterJitter1}). Following the void creation the agents are now impacted by a second concave reduction from the newly created void and the anomaly on the perimeter edge `snaps' back closing the void~(Figure~\ref{fig:OuterPerimeterJitter2}). This process repeats itself creating an instability in the number of perimeter agents which is highlighted in~Figure~\ref{concave:BaselineConcavePerimeter} as the erratic change in the number of agents.

@@

Figure~\ref{concave:BaselineConcaveEffectPath1} shows the paths of the agents in the swarm with concave reduction (red) and the baseline (black). The paths of the agents are initially very similar as the swarm expands but as the \textit{interaction vector magnitudes} rise and the swarm stabilises the effect of the \textit{concave reduction vectors} start to noticeably influence the swarm structure. The most noticeable effect is on the perimeter where the agents have expanded then instead of stabalising to relatively stable fixed position there is drifting effect occurring due to the imbalance of the initial deployment structure (more anomalies on one side). The swarm also becomes more `rounded' in appearance.

@@

Figure~\ref{concave:BaselineConcaveEffectPath2} shows a more detailed view of the structure within the swarm. The baseline (black) paths show the swarm expanding and then settling to a hexagonal pattern which appears to oscillate slightly (jitter) when the swarm has reached its optimum distribution. The concave reduction swarm agents paths (red) show the swarm expanding in a similar way but once fully expanded the concave reduction causes the swarm to move with a slight directional bias.

@@

To reduce the `snapping' effect (shown in Figures~\ref{fig:OuterPerimeterJitter1} and \ref{fig:OuterPerimeterJitter2}), the field effect parameters can be adjusted to create a greater tolerance in the agent interactions. This can be achieved by either reducing the agents repulsion field and maintaining the neighbour field effect or increasing the neighbour field effect and maintaining the repulsion field effect. These changes affect the structure of the swarm but ensure that the agents to stay within the cohesion field when moving to implement the concave reduction the agents are therefore `held' by the cohesion field effect preventing the `snap'. These changes in the field effect can be shown experimentally using the parameters in~Table~\ref{tab:BaselineConcaveReduction2}.

@@

Figure~\ref{concave:BaselineConcaveEffectDist8060} shows a comparison of the agent movements based on distance for the revised field effects. The graph shows that the swarm settles to a distance that is closer due to the reduced repulsion field. The graph also shows that the concave reduction still induces additional jitter as the variance is still greater than the baseline but due to the reduced snapping the perimeter agent count shows greater stability~(\ref{concave:BaselineConcavePerimeter8060}).

@@

Figure~\ref{concave:BaselineConcaveEffectMag8060} shows that the \textit{inter-agent vector magnitude} is reduced due to the reduced cohesion from the agent proximity and the \textit{concave reduction vector magnitude} impact is reduced due to less anomalies occurring on the perimeter.
%BASELINE-COMPRESS-MAG-8060.py
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{figures/BaselineConcaveEffectMag8060}
\end{center}
\caption{Baseline/Concave effect magnitude (80/60)\label{concave:BaselineConcaveEffectMag8060}}
\end{figure}

Figure~\ref{concave:BaselineConcavePerimeter8060} shows the changes in the perimeter size from the baseline and the concave reduction swarms. The concave reduction still creates an erratic perimeter count but the variation is reduced. On aggregate for the run the concave reduction has reduced the perimeter size. The revised effects also reduce the snapping effect and the swarm has an improved structure. The concave reduction creates a perimeter which fluctuates between 45-50 agents where as the baseline swarm settles to 49 agents.

@@

Figure~\ref{concave:BaselineConcaveEffectPath80601} shows the paths of the agents in the swarm with concave reduction (red) and without (black). The paths are similar as the swarm expands due to the \textit{repulsion vectors} being large and masking the \textit{concave reduction vectors} due to the resultant \textit{movement vector} directing the agents in a similar direction. Once the swarm has expanded the \textit{concave reduction vectors} create a more spherical appearance to the swarm. The \textit{concave reduction vector magnitudes} are significantly large enough now to create a slight directional bias as shown in both~Figures~\ref{concave:BaselineConcaveEffectPath80601} and~\ref{concave:BaselineConcaveEffectPath80602}. This effect is the result of the \textit{concave reduction vectors} pushing the anomalies on the left of the swarm resulting in the swarm moving slightly to the right due to the swarm having fewer anomalies on the opposite side of the swarm. The anomalies can be seen in~Figure~\ref{concave:BaselineConcaveEffectPath80601}.

@@

With the tolerance levels set appropriately the impact of the concave reduction is to reduce the number of agents being identified as coordinators~(Figure~\ref{concave:mobileSwarm1}). This reduction in perimeter size is due to a reduction in the number of anomalies in the swarm. The effect of the concave reduction on a convex (outer) perimeter is limited when a swarm is deployed in an almost circular manner as there is limited space for optimisation. The effect is more pronounced when a swarm is `malformed' with large anomalies producing concave edges.

@@

\section{Application of concave reduction on concave perimeters (voids)}\label{sec:ApplicationConcavePerimeters}
Just as a gap on a convex perimeter is affected by concave reduction so is the perimeter of a concave perimeter (a void). The effect on a concave perimeter is more pronounced due to there being a higher ratio of concave anomalies. It is possible to have no concave gaps on a convex perimeter (a circular swarm) but a concave perimeter (void) always has concave anomalies. This characteristic allows voids to be controlled. To test the effect of concave reduction on void removal a baseline must be established~(Figure~\ref{tab:BaselineConcaveReduction3}).

@@

Figures~\ref{fig:BaselineEndPoint4560} and \ref{fig:ConcaveEndPoint4560} shows the end points for the simulations for the concave reduction experiments. Figure~\ref{fig:BaselineEndPoint4560} shows the end point for the baseline. It shows that at the end of the simulation the void within the swarm persists. This is due to the distribution of the agents being optimal for the given simulation parameters and the structures within the swarm being `stable'. Figure~\ref{fig:ConcaveEndPoint4560} shows the end point for the simulation with concave reduction enabled using the same swarm and configuration parameters. The concave reduction algorithm has removed the void from the swarm completely and the outer perimeter has been `smoothed'.

@@

Figures \ref{voids:ConcavePerimeter4560-DIST}, \ref{voids:ConcavePerimeter4560-DIST-2}, \ref{voids:ConcavePerimeter4560-MAG}, \ref{voids:ConcavePerimeter4560-MAG-2} show the effect the concave reduction has on the inter-agent distances and \textit{inter-agent vector magnitudes} for the simulation.

>>>>

\begin{abstract}
%\boldmath
Swarms consist of many agents and this is exploited to make them tolerant of failures. Failures occur for may reasons; for example: resource exhaustion, component failure, or a disruption from an external event. The loss of agents reduces the size of a swarm and changes the structure in places where the agents are lost. A swarm's structure can also be irregular from an initial deployment or due to an obstacle in its path. These changes in the structure and/or size do not stop a swarm from functioning but can affect the outcome of a task to which the swarm is being applied. The principle behind concave reduction is to counter the effect of agent loss or irregular structures within a swarm. Depending upon the type of failure or disruption sections of the swarm can develop holes (areas that should contain agents) or the overall shape may become distorted. Concave reduction is a technique where each agent identifies its position in relation to its neighbours and if the agent meets specific criteria it will attempt to move in such a way as to remove the gap between two of its neighbours. These changes in direction of multiple agents have a restructuring effect which results in a more uniform distribution of agents and reduces the perimeter size for the number of agents in the swarm.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

\section{Swarm coordination - concave reduction}\label{chapter:ConcaveReduction}
When cohesion and repulsion (\textit{inter-agent vectors}) are the only field effects creating a swarming effect the number of stable structures that can develop is limited. These structures are either straight edges or partial lattices. Partial lattices can create concave anomalies (dents) and convex anomalies (peaks) in a perimeter~(Fig.~\ref{concave:SwarmStableShape1}). An anomaly is any construct within the swarm that is not a hexagonal lattice or an agent distribution that causes the swarm to deviate from having only convex or flat edges.

@@

When discussing concave reduction the concave angle of interest is the relationship between an agent and two of its adjacent neighbours such that the neighbours have no visibility of each other and the neighbour-agent-neighbour relationship produces an angle~$< 180^\circ$. This is unrelated to the aggregate angle used to identify the perimeter type that the agent is a member of. Concave reduction therefore affects both inner~(Fig.~\ref{fig:InnerPerimeter1}) and outer~(Fig.~\ref{fig:OuterPerimeter1}) perimeter agents. Fig.~\ref{fig:OuterPerimeter1} and \ref{fig:InnerPerimeter1} show candidate agents identified by this process in black.

@@

The purpose of concave reduction is to alter the overall structure of the swarm by replacing the calculated \textit{movement-direction vector} on anomalous perimeter agents. These perimeter based changes cause a cascading movement within the swarm as non-perimeter based agent's \textit{interaction vectors} attempting to move towards a position of equilibrium to their neighbours. The movement of the concave-perimeter~(internal void) agents creates a tendency for the void to be removed~(Fig.~\ref{fig:InnerPerimeter2}~and~\ref{fig:InnerPerimeter3}) by percolating the void out of the swarm. On the outer perimeter deformities (indents) are straightened~(Fig.~\ref{fig:OuterPerimeter3}) and eventually a perimeter agent and its neighbours will tend towards creating a convex edge.

@@

\subsection{Full-perimeter coordinator detection}\label{sec:PerimeterAgentDetection} 
The process of detecting a perimeter agent involves several steps. Primarily it involves identifying agents that are not surrounded by interconnected neighbours~(Fig.~\ref{reduced:Perimeter1}). The processes are discussed in more detail below:

@@

The first check is a short circuit count of the number of neighbours. If an agent has $\leq 4$ neighbours and the swarm parameters are for a hexagonal swarm configuration then the agent is a perimeter agent and therefore a coordinator. This identification process is shown in~Fig.~\ref{algo:SmallNeighbourCount}.

@@

If the short circuit neighbour count check fails then further conditions must be checked. The additional checks require a dictionary of agent/angular values to be generated~\eqref{eq:ConcaveVoidPhysics1}. Fig.~\ref{algo:getNeighbourAngles} shows how the dictionary of agent/angle values are produced.

@@

Fig. \ref{algo:getNeighbourAngles} shows the logic to produce the sorted neighbour/angle dictionary. $sort(.)$ sorts the dictionary set generated in~\eqref{eq:ConcaveVoidPhysics1} by ascending angle to produce a dictionary of neighbour agents with their relative angle to the first detected neighbour as shown in~Fig.~\ref{reduced:Perimeter1}.

@@

Using the angular dictionary from Fig.~\ref{algo:getNeighbourAngles} it is possible to identify if adjacent neighbours can detect each other. If the agent is within all of its neighbours boundary then it is not on the perimeter~(Fig.~\ref{algo:CheckVisibility}).

@@

The full visibility check~(Fig.~\ref{algo:CheckVisibility}) takes each pair in turn and using the cosine rule~\eqref{eq:CosineRule1} checks for a `visibility gap'.

@@

There is one exception to the neighbour visibility check. Due to compression of the swarm, which can be caused by an initial deployment configuration or when an obstacle is in the path of the swarm, the agent's neighbours are able to `see' other neighbours but a pair of sequential neighbours could create an angle $> 180^\circ$ as shown in~Fig.~\ref{reduced:Perimeter2}. In this case the agent is on the outside of the enclosed neighbour space, the agent is therefore on a perimeter.

@@

This phenomenon requires the detection process to identify the angles that all the agents create from an arbitrary point, in this case the first identified neighbour. The neighbours must then be checked to determine if any neighbour pair create an angle $> 180^\circ$~(Fig.~\ref{reduced:Perimeter1}~Fig.~\ref{algo:CheckConvex}). This process includes a short circuit in that as soon as a gap is found the check terminates.

@@

The complete perimeter detection algorithm using this `cyclic-angular-neighbour-check' methodology is show in~Fig.~\ref{algo:CheckPerimeter}. The first part of the algorithm is the short circuit check~(\textit{SmallNeighbourCount(b)}). This is only followed by the visibility check (\textit{CheckVisibility(b,angles)}) if the short circuit fails. If the visibility check is confirmed, which is another short circuit method, the convex check is carried out~(\textit{CheckConvex(angles)}). This sequence ensures the minimum computational overhead for the agent check.

@@

In a static swarm, where there are no \textit{destination vectors}, concave reduction will result in a restructuring motion that creates a more `rounded' swarm~(Fig.~\ref{fig:OuterPerimeter3}). Concave reduction also creates a surround effect as it removes voids from a swarm~(Section~\ref{voids:ObjectSurrounding}). When concave reduction is applied to a mobile swarm the void closing effect can be used to improve reconnaissance coverage as a swarm passes around obstacles~(Section~\ref{concave:mobileSwarm1}).

@@

The implementation of concave reduction identification is incorporated into the perimeter detection algorithm as shown in Fig.~\ref{algo:checkVisibility2}.\\

@@

\section{Concave reduction agent movement}\label{concave:AgentMovement}
Adding a further characteristic to the motion of a swarm necessitates a revision to the existing agent model~\eqref{eq:InterAgentMovement2}. With concave reduction this revision is based upon the identification of the concave perimeter edges as shown in~(Fig.~\ref{concave:VoidConcave1}) $(b^{'},b,b^{''})$.

@@

When an agent is identified as being a component of a concave characteristic~(Fig.~\ref{algo:checkVisibility2}) the normal \textit{movement-direction vector}~\eqref{eq:InterAgentMovement2} is replaced by a \textit{concave reduction vector}. This new vector causes the agent to move in a direction that will reduce or remove a concave edge by moving the agent towards the identified gap. The effect of this will be to either straighten an outer perimeter or reduce/remove a void. This change in direction increases the distance and magnitude variances (jitter) due to the changes it will induce in the neighbours \textit{interaction vectors}.

Fig.~\ref{fig:InterAgentEffect1} and \ref{fig:InterAgentEffect2} shows this effect in more detail. Fig.~\ref{fig:InterAgentEffect1}~shows the initial positions of the agents before the concave reduction affects the agent. Fig.~\ref{fig:InterAgentEffect2}~shows the positive and negative effect on the inter-agent distances that the movement creates. The aggregate change is an increase in the inter-agent distances and an increase in the resultant magnitude effect.

@@

\subsection{Perimeter Exceptions}\label{concave:Exceptions}
When a swarm is forming or disrupted through either agent failures, catastrophic events or interactions with obstacles sections of the swarm may become isolated. When this occurs the sections can be considered as being a separate sub-swarms. It is also possible for sections of the swarm to remain connected through tenuous links of either 1 or 2 agents as shown in Fig.~\ref{fig:Connector1} and \ref{fig:Connector2}. In the case of a link connection~(Fig.~\ref{fig:Connector1}) concave reduction can be applied as normal. In the case of a `bridge' link (Fig.~\ref{fig:Connector2}) there is an issue in that there is more than one concave gap. As there is no communications infrastructure in swarm models there are limited options for handling this situation. One option is allow the swarming algorithm to apply as usual and no concave reduction takes place; the reasoning being that the overall structure of the swarm is unknown therefore the movement cannot be optimised. An alternative would be to select either the largest or the smallest gap; again as the structure of the swarm is unknown this may not be beneficial. A third option would be to select one of the gaps in some way (randomly, first seen, last seen) and implement the concave reduction at that point. The approach taken in this paper is to select the first gap that is detected. Selecting the first gap reduces the computational overhead of the algorithm.

@@

Fig.~\ref{concave:BaselineConcaveEffectDist} shows the changes in the inter-agent distances that result from the concave reduction. The experiment demonstrates that with concave reduction the average distance increases and the variance of the distances increases. This is due to the `pulling' effect of the \textit{concave reduction vector} on the perimeter. The algorithm distorts the distribution of the agents as discussed in section~\ref{concave:AgentMovement}. The initial expansion of the swarm is similar for the baseline and the concave reduction however at 9 seconds into the simulation the \textit{concave reduction vectors} affects the swarm sufficiently to prevent the average distance reducing to the same level as the baseline. Once the swarm has stabilised with the concave reduction enabled the average distance and variance remain constant but above the baseline.

@@

Fig.~\ref{concave:BaselineConcaveEffectMag} shows the change in the \textit{inter-agent vector magnitudes} that the concave reduction introduces. Both the average magnitude increases and the variance. The magnitude increases as the \textit{concave reduction vector} has caused the agents to move outwards increasing the area of the swarm. The increase in cohesion is a direct result of the increased distance. The field effects still intersect but due to the average distance of the agents being further apart the repulsion is reduced. The agents being further apart causes the increase in cohesion as the algorithm attempts to prevents the swarm from breaking up. The variance increase is caused by the concave reduction effect moving selected agents into less optimal positions.

@@

Fig.~\ref{concave:BaselineConcavePerimeter} shows the effect on the number of perimeter agents. Initially the perimeter size is minimally affected due to the swarms compression however after the expansion phase of the swarm (8 seconds) more perimeter agents are identified as concave edges are formed. As these concave edges are `straightened' they create further anomalies that ripple through the swarm. This rippling is identified by the erratic changes in the number of perimeter agents. After approximately 30 seconds the swarm has been forced into a less angular structure with curved edges and the number of perimeter agents falls below the minimum of the baseline. The shape of the swarm is still undergoing change following this and the perimeter count continues to fluctuate.

@@

The effect on the structure of the swarm caused by the concave perimeter agents moving towards the gap agents is to pull the internal agents forward. This pulling causes the swarm to develop a more rounded structure but the effect also have a negative impact. If the tolerance of the agent's movements (the difference between the agent ranges for repulsion and cohesion) is too small then the distortion effect of the concave reduction can create additional voids. The perimeter agents will then move so as to remove the defect~(Fig.~\ref{fig:OuterPerimeterJitter1}). Following the void creation the agents are now impacted by a second concave reduction from the newly created void and the anomaly on the perimeter edge `snaps' back closing the void~(Fig.~\ref{fig:OuterPerimeterJitter2}). This process repeats itself creating an instability in the number of perimeter agents which is highlighted in~Fig.~\ref{concave:BaselineConcavePerimeter} as the erratic change in the number of agents.

@@

Fig.~\ref{concave:BaselineConcaveEffectPath1} shows the paths of the agents in the swarm with concave reduction (red) and the baseline (black). The paths of the agents are initially very similar as the swarm expands but as the \textit{interaction vector magnitudes} rise and the swarm stabilises the effect of the \textit{concave reduction vectors} start to noticeably influence the swarm structure. The most noticeable effect is on the perimeter where the agents have expanded then instead of stabalising to relatively stable fixed position there is drifting effect occurring due to the imbalance of the initial deployment structure (more anomalies on one side). The swarm also becomes more `rounded' in appearance.

@@

Fig.~\ref{concave:BaselineConcaveEffectPath2} shows a more detailed view of the structure within the swarm. The baseline (black) paths show the swarm expanding and then settling to a hexagonal pattern which appears to oscillate slightly (jitter) when the swarm has reached its optimum distribution. The concave reduction swarm agents paths (red) show the swarm expanding in a similar way but once fully expanded the concave reduction causes the swarm to move with a slight directional bias.

@@

To reduce the `snapping' effect (shown in Fig.~\ref{fig:OuterPerimeterJitter1} and \ref{fig:OuterPerimeterJitter2}), the field effect parameters can be adjusted to create a greater tolerance in the agent interactions. This can be achieved by either reducing the agents repulsion field and maintaining the neighbour field effect or increasing the neighbour field effect and maintaining the repulsion field effect. These changes affect the structure of the swarm but ensure that the agents to stay within the cohesion field when moving to implement the concave reduction the agents are therefore `held' by the cohesion field effect preventing the `snap'. These changes in the field effect can be shown experimentally using the parameters in~Table~\ref{tab:BaselineConcaveReduction2}.

@@

Fig.~\ref{concave:BaselineConcaveEffectDist8060} shows a comparison of the agent movements based on distance for the revised field effects. The graph shows that the swarm settles to a distance that is closer due to the reduced repulsion field. The graph also shows that the concave reduction still induces additional jitter as the variance is still greater than the baseline but due to the reduced snapping the perimeter agent count shows greater stability~(\ref{concave:BaselineConcavePerimeter8060}).

@@

Fig.~\ref{concave:BaselineConcaveEffectMag8060} shows that the \textit{inter-agent vector magnitude} is reduced due to the reduced cohesion from the agent proximity and the \textit{concave reduction vector magnitude} impact is reduced due to less anomalies occurring on the perimeter.
%BASELINE-COMPRESS-MAG-8060.py
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{figures/BaselineConcaveEffectMag8060}
\end{center}
\caption{Baseline/Concave effect magnitude (80/60)\label{concave:BaselineConcaveEffectMag8060}}
\end{figure}

Fig.~\ref{concave:BaselineConcavePerimeter8060} shows the changes in the perimeter size from the baseline and the concave reduction swarms. The concave reduction still creates an erratic perimeter count but the variation is reduced. On aggregate for the run the concave reduction has reduced the perimeter size. The revised effects also reduce the snapping effect and the swarm has an improved structure. The concave reduction creates a perimeter which fluctuates between 45-50 agents where as the baseline swarm settles to 49 agents.

@@

Fig.~\ref{concave:BaselineConcaveEffectPath80601} shows the paths of the agents in the swarm with concave reduction (red) and without (black). The paths are similar as the swarm expands due to the \textit{repulsion vectors} being large and masking the \textit{concave reduction vectors} due to the resultant \textit{movement vector} directing the agents in a similar direction. Once the swarm has expanded the \textit{concave reduction vectors} create a more spherical appearance to the swarm. The \textit{concave reduction vector magnitudes} are significantly large enough now to create a slight directional bias as shown in both~Fig.~\ref{concave:BaselineConcaveEffectPath80601} and~\ref{concave:BaselineConcaveEffectPath80602}. This effect is the result of the \textit{concave reduction vectors} pushing the anomalies on the left of the swarm resulting in the swarm moving slightly to the right due to the swarm having fewer anomalies on the opposite side of the swarm. The anomalies can be seen in~Fig.~\ref{concave:BaselineConcaveEffectPath80601}.

@@

With the tolerance levels set appropriately the impact of the concave reduction is to reduce the number of agents being identified as coordinators~(Fig.~\ref{concave:mobileSwarm1}). This reduction in perimeter size is due to a reduction in the number of anomalies in the swarm. The effect of the concave reduction on a convex (outer) perimeter is limited when a swarm is deployed in an almost circular manner as there is limited space for optimisation. The effect is more pronounced when a swarm is `malformed' with large anomalies producing concave edges.

@@

\section{Application of concave reduction on concave perimeters (voids)}\label{sec:ApplicationConcavePerimeters}
Just as a gap on a convex perimeter is affected by concave reduction so is the perimeter of a concave perimeter (a void). The effect on a concave perimeter is more pronounced due to there being a higher ratio of concave anomalies. It is possible to have no concave gaps on a convex perimeter (a circular swarm) but a concave perimeter (void) always has concave anomalies. This characteristic allows voids to be controlled. To test the effect of concave reduction on void removal a baseline must be established~(Fig.~\ref{tab:BaselineConcaveReduction3}).

@@

Fig.~\ref{fig:BaselineEndPoint4560} and \ref{fig:ConcaveEndPoint4560} shows the end points for the simulations for the concave reduction experiments. Fig.~\ref{fig:BaselineEndPoint4560} shows the end point for the baseline. It shows that at the end of the simulation the void within the swarm persists. This is due to the distribution of the agents being optimal for the given simulation parameters and the structures within the swarm being `stable'. Fig.~\ref{fig:ConcaveEndPoint4560} shows the end point for the simulation with concave reduction enabled using the same swarm and configuration parameters. The concave reduction algorithm has removed the void from the swarm completely and the outer perimeter has been `smoothed'.

@@

Fig. \ref{voids:ConcavePerimeter4560-DIST}, \ref{voids:ConcavePerimeter4560-DIST-2}, \ref{voids:ConcavePerimeter4560-MAG}, \ref{voids:ConcavePerimeter4560-MAG-2} show the effect the concave reduction has on the inter-agent distances and \textit{inter-agent vector magnitudes} for the simulation.

Fig.~\ref{voids:ConcavePerimeter4560-DIST} shows that the process of reducing the void increases the average distance of the agents. This is caused by the concave agents `pulling' away from their neighbours. The baseline experiment shows a limited change in the variance (jitter) due to the swarm being close to stable even though there is a void present.
